# Day 2 — Linear regression with interaction effects

The dependent variable is **social trust (`ppltrst`)**. Predictors come from media use and demographics in the ESS subset.

**Model notation recap**

- Baseline linear model: \(Y_i = \beta_0 + \mathbf{x}_i^\top \boldsymbol\beta + \varepsilon_i\), \( \varepsilon_i \sim \text{i.i.d. }(0, \sigma^2)\).
- Binary–binary interaction (e.g., gender × urban): \(Y_i = \beta_0 + \beta_1 \text{Female}_i + \beta_2 \text{Urban}_i + \beta_3 (\text{Female}_i \times \text{Urban}_i) + \dots\).  
  - \(\beta_3\) is the *difference-in-differences*: the extra gap between women and men *when Urban = 1* minus the gap when Urban = 0.
- Binary–continuous interaction (gender × age): slope for age becomes \(\beta_{\text{age}} + \beta_{\text{age} \times \text{female}} \cdot \text{Female}_i\); draw ribbons to see how slopes differ across groups.
- Three-way interaction (gender × age × country): the age slope is country- and gender-specific: \(\partial Y / \partial \text{age} = \beta_{\text{age}} + \beta_{\text{age}\times g} g + \beta_{\text{age}\times c} c + \beta_{\text{age}\times g \times c} g c\).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4)
Sys.setenv(OMP_NUM_THREADS = 1,
           OPENBLAS_NUM_THREADS = 1,
           MKL_NUM_THREADS = 1,
           VECLIB_MAXIMUM_THREADS = 1,
           OMP_WAIT_POLICY = "PASSIVE",
           OMP_DYNAMIC = "FALSE",
           OMP_THREAD_LIMIT = 1,
           KMP_DUPLICATE_LIB_OK = "TRUE",
           KMP_INIT_AT_FORK = "FALSE")
```

```{r fits}
library(dplyr)
library(ggplot2)
library(broom)
library(purrr)
library(tidyr)

source("R/clean_ess.R")

ess <- clean_ess()

# Fit once and reuse
m0 <- lm(ppltrst ~ agea + gender + news_days + country, data = ess)
m1 <- lm(ppltrst ~ gender * urban + agea + news_days + country, data = ess)
m2 <- lm(ppltrst ~ gender * agea + news_days + country, data = ess)
m3 <- lm(ppltrst ~ agea * news_days + gender + country, data = ess)
m4 <- lm(ppltrst ~ gender * agea * country + news_days, data = ess)

# Predicted values for plots (simple grids)
nd1 <- expand.grid(
  urban = c("Urban", "Non-urban"),
  gender = c("Male", "Female"),
  agea = mean(ess$agea, na.rm = TRUE),
  news_days = mean(ess$news_days, na.rm = TRUE),
  country = "GB"
)
pred1 <- predict(m1, newdata = nd1, se.fit = TRUE)
nd1$fit <- pred1$fit
nd1$lo <- pred1$fit - 1.96 * pred1$se.fit
nd1$hi <- pred1$fit + 1.96 * pred1$se.fit
int_plot1 <- ggplot(nd1, aes(x = urban, y = fit, fill = gender)) +
  geom_col(position = position_dodge(width = 0.6), width = 0.5) +
  geom_errorbar(aes(ymin = lo, ymax = hi), position = position_dodge(width = 0.6), width = 0.2) +
  labs(y = "Predicted social trust", x = "Residential area") +
  theme_minimal()

age_seq <- seq(min(ess$agea, na.rm = TRUE), max(ess$agea, na.rm = TRUE), length.out = 60)
nd2 <- expand.grid(agea = age_seq, gender = c("Male", "Female"),
                   news_days = mean(ess$news_days, na.rm = TRUE),
                   country = "GB")
pred2 <- predict(m2, newdata = nd2, se.fit = TRUE)
nd2$fit <- pred2$fit
nd2$lo <- pred2$fit - 1.96 * pred2$se.fit
nd2$hi <- pred2$fit + 1.96 * pred2$se.fit
int_plot2 <- ggplot(nd2, aes(x = agea, y = fit, color = gender)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = gender), alpha = 0.15, color = NA) +
  labs(y = "Predicted social trust", x = "Age") +
  theme_minimal()

news_seq <- seq(min(ess$news_days, na.rm = TRUE), max(ess$news_days, na.rm = TRUE), length.out = 40)
nd3 <- expand.grid(news_days = news_seq,
                   agea = quantile(ess$agea, c(.2, .5, .8), na.rm = TRUE),
                   gender = "Male",
                   country = "GB")
pred3 <- predict(m3, newdata = nd3, se.fit = TRUE)
nd3$fit <- pred3$fit
nd3$lo <- pred3$fit - 1.96 * pred3$se.fit
nd3$hi <- pred3$fit + 1.96 * pred3$se.fit
int_plot3 <- ggplot(nd3, aes(x = news_days, y = fit, color = factor(agea))) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = factor(agea)), alpha = 0.12, color = NA) +
  labs(y = "Predicted social trust", x = "News days per week", color = "Age quantile") +
  theme_minimal()

nd4 <- expand.grid(agea = age_seq,
                   gender = c("Male","Female"),
                   country = c("GB","DE","FR"),
                   news_days = mean(ess$news_days, na.rm = TRUE))
pred4 <- predict(m4, newdata = nd4, se.fit = TRUE)
nd4$fit <- pred4$fit
nd4$lo <- pred4$fit - 1.96 * pred4$se.fit
nd4$hi <- pred4$fit + 1.96 * pred4$se.fit
int_plot4 <- ggplot(nd4, aes(x = agea, y = fit, color = gender)) +
  geom_line() +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = gender), alpha = 0.12, color = NA) +
  facet_wrap(~ country) +
  labs(y = "Predicted social trust", x = "Age") +
  theme_minimal()

# OLS intuition demo (interactive)
ess_small <- ess |> select(ppltrst, agea) |> drop_na() |> slice_sample(n = 600)
fit_simple <- lm(ppltrst ~ agea, data = ess_small)

base_resid <- ggplot(ess_small, aes(x = agea, y = ppltrst)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = coef(fit_simple)[2], intercept = coef(fit_simple)[1], color = "#4C78A8", size = 1.1) +
  geom_segment(aes(xend = agea, yend = fitted(fit_simple)), alpha = 0.25, color = "#9ecae1") +
  labs(x = "Age", y = "Trust (0-10)", title = "OLS fit with residuals") +
  theme_minimal()

slope_grid <- seq(coef(fit_simple)[2] - 0.08, coef(fit_simple)[2] + 0.08, length.out = 30)
anim_df <- map_dfr(slope_grid, ~{
  pred <- coef(fit_simple)[1] + .x * ess_small$agea
  tibble(agea = ess_small$agea,
         ppltrst = ess_small$ppltrst,
         slope = sprintf("%.3f", .x),
         pred = pred,
         resid = ppltrst - pred)
})
anim_plot <- ggplot(anim_df, aes(x = agea, y = ppltrst, frame = slope)) +
  geom_point(alpha = 0.35) +
  geom_abline(aes(slope = as.numeric(slope), intercept = coef(fit_simple)[1]), color = "#4C78A8", size = 1) +
  labs(x = "Age", y = "Trust (0-10)", title = "Searching for the best-fit slope") +
  theme_minimal()
```

## OLS intuition: best-fit line and residuals {.tabset}

#### Output

```{r ols-static, echo=FALSE}
base_resid
```

```{r ols-intuition-out, echo=TRUE, eval=TRUE}
# To see the moving line locally:
# library(plotly)
# subplot(
#   ggplotly(base_resid),
#   ggplotly(anim_plot) %>% animation_opts(frame = 250, transition = 0, redraw = FALSE),
#   nrows = 2, heights = c(0.55, 0.45), margin = 0.05
# )
```

Interpretation: the static panel shows residuals; if you run the optional plotly code locally, the moving line illustrates how residuals shrink as the slope approaches the OLS solution.

## 1. Baseline linear model {.tabset}

#### Code

```{r m0-code, eval=FALSE}
m0 <- lm(ppltrst ~ agea + gender + news_days + country, data = ess)
broom::tidy(m0)
```

#### Output

```{r m0-out, echo=FALSE}
broom::tidy(m0)
```
Interpretation: Trust decreases slightly with age and differs by country and gender; focus on sign and magnitude of the coefficients rather than raw p-values when discussing effect sizes.

## 2. Binary × Binary interaction (gender × urban) {.tabset}

#### Code

```{r m1-code, eval=FALSE}
m1 <- lm(ppltrst ~ gender * urban + agea + news_days + country, data = ess)
int_plot1 <- ggplot(nd1, aes(x = urban, y = fit, fill = gender)) +
  geom_col(position = position_dodge(width = 0.6), width = 0.5) +
  labs(y = "Predicted social trust", x = "Residential area") +
  theme_minimal()
```

#### Output

```{r m1-out, echo=FALSE}
broom::tidy(m1)
int_plot1
```
Interpretation: The urban–rural trust gap is small; note whether the CI bars for Male vs Female overlap. If they do, the moderation by gender is likely negligible.

Interpretation focus: Does the urban–rural gap differ by gender?

## 3. Binary × Continuous interaction (gender × age) {.tabset}

#### Code

```{r m2-code, eval=FALSE}
m2 <- lm(ppltrst ~ gender * agea + news_days + country, data = ess)
int_plot2 <- ggplot(nd2, aes(x = agea, y = fit, color = gender)) +
  geom_line(size = 1) +
  labs(y = "Predicted social trust", x = "Age") +
  theme_minimal()
```

#### Output

```{r m2-out, echo=FALSE}
broom::tidy(m2)
int_plot2
```
Interpretation: Slopes by age differ by gender; parallel ribbons would imply no interaction. Diverging ribbons indicate the age effect depends on gender.

Key idea: slopes for age are estimated separately for men and women.

## 4. Continuous × Continuous interaction (age × news consumption) {.tabset}

#### Code

```{r m3-code, eval=FALSE}
m3 <- lm(ppltrst ~ agea * news_days + gender + country, data = ess)
int_plot3 <- ggplot(nd3, aes(x = news_days, y = fit, color = factor(agea))) +
  geom_line(size = 1) +
  labs(y = "Predicted social trust", x = "News days per week", color = "Age quantile") +
  theme_minimal()
```

#### Output

```{r m3-out, echo=FALSE}
broom::tidy(m3)
int_plot3
```
Interpretation: Check whether the news-consumption slope changes across age quantiles; overlapping ribbons mean little moderation, separated ribbons suggest stronger news effects at certain ages.

Discuss whether news exposure moderates the age–trust relationship.

## 5. Three-way interaction (gender × age × country) {.tabset}

#### Code

```{r m4-code, eval=FALSE}
m4 <- lm(ppltrst ~ gender * agea * country + news_days, data = ess)
int_plot4 <- ggplot(nd4, aes(x = agea, y = fit, color = gender)) +
  geom_line() +
  facet_wrap(~ country) +
  labs(y = "Predicted social trust", x = "Age") +
  theme_minimal()
```

#### Output

```{r m4-out, echo=FALSE, fig.height=5}
broom::tidy(m4)
int_plot4
```
Interpretation: Three-way plots show country-specific age slopes by gender; look for countries where ribbons separate widely—that’s where the interaction is substantive.

Strategy: interpret pairwise contrasts within each country before comparing across countries.

## 6. Model comparison and diagnostics

````r
broom::glance(m0, m1, m2, m3, m4)
# quick residual check
par(mfrow = c(2,2)); plot(m3)
````

## Problem set — Interaction lab

1. Refit `m1` but swap `urban` with a binary indicator for *high education* (e.g., `eduyrs >= 15`). Interpret the gender gap at low vs high education.
2. Build a model with `ppltrst ~ news_days * country + agea + gender`. Compute marginal effects of `news_days` within each country.
3. Add a three-way term `gender * urban * country`. Plot predicted trust for all six gender-by-urban-by-country profiles.
4. Briefly report which interaction improves fit (compare adjusted R² and AIC) and whether the effect is substantively meaningful.

Use `marginaleffects::plot_slopes()` and `plot_predictions()` to visualise interactions instead of only staring at coefficients.
