# Survey weighting — why and how

Survey data are collected with unequal selection probabilities. Inference should reflect the design to avoid biased point estimates and standard errors.

- **Notation:** Let \(w_i = 1/\pi_i\) be the design (inverse-probability) weight. For a finite population mean \(\bar{Y} = \sum_i w_i Y_i / \sum_i w_i\). For regression, weighted likelihoods re-scale each case by \(w_i\).
- **ESS fields used:** `pweight` (post-stratification weight), `psu` (primary sampling unit), `stratum` (strata). We set `options(survey.lonely.psu = "adjust")` to stabilize single-PSU strata.

### What these design variables mean (plain language)

- **`pweight` (post-stratification weight):** Adjusts for unequal inclusion probabilities *and* aligns the achieved sample with known population margins (e.g., age × gender × region). Large `pweight` values upweight under-represented respondents; small values downweight over-represented ones.
- **`psu` (primary sampling unit):** The first cluster stage of selection (e.g., municipalities or postcode sectors). Respondents inside the same PSU share fieldwork and selection features, so their responses are correlated.
- **`stratum` (strata):** Mutually exclusive groups within which PSUs were sampled (e.g., by region × urbanicity). Stratification improves precision; variance estimation must respect it.
- **Design degrees of freedom:** With clustering and stratification, the effective df are closer to the number of PSUs minus strata, not the raw respondent count—hence the importance of design-aware SEs.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4)
Sys.setenv(OMP_NUM_THREADS = 1,
           OPENBLAS_NUM_THREADS = 1,
           MKL_NUM_THREADS = 1,
           VECLIB_MAXIMUM_THREADS = 1)
```

```{r packages}
library(dplyr)
library(ggplot2)
library(tidyr)
library(survey)
library(broom)

# Synthetic microdata to illustrate weighting without depending on raw ESS file quirks
set.seed(42)
n_psu <- 120
n_by_psu <- sample(8:18, n_psu, replace = TRUE)

demo_psu <- tibble(
  psu = 1:n_psu,
  stratum = sample(1:20, n_psu, replace = TRUE),
  country = sample(c("GB", "DE", "FR"), n_psu, replace = TRUE, prob = c(.4, .35, .25)),
  pweight = runif(n_psu, 0.5, 3),
  n = n_by_psu
)

ess_w <- demo_psu |>
  uncount(n) |>
  mutate(
    agea = round(rnorm(n(), 50, 15)),
    gender = sample(c("Male", "Female"), n(), replace = TRUE),
    # Generate outcome with country- and age-dependent probability
    linpred = -1 + 0.6 * (country == "GB") + 0.3 * (country == "FR") +
              0.01 * (agea - 50) + 0.4 * (gender == "Female"),
    news_regular = rbinom(n(), 1, plogis(linpred))
  ) |>
  select(psu, stratum, country, pweight, agea, gender, news_regular) |>
  mutate(country = factor(country),
         gender = factor(gender))

options(survey.lonely.psu = "adjust")
des <- svydesign(ids = ~psu, strata = ~stratum, weights = ~pweight,
                 data = ess_w, nest = TRUE)

svy_n <- nrow(des$variables)
```

## 1. Weighted descriptive estimates

Sample used after complete-case filtering: `r svy_n` respondents.

```{r weighted-mean}
svymean(~news_regular, des)
```

- The point estimate is the **design-weighted mean**; SEs account for clustering and stratification.
- **Design effect** (`svymean(..., deff=TRUE)`) tells how much variance inflation comes from the design versus SRS.

### Country-weighted proportions

```{r weighted-country, echo=FALSE}
pr_cty <- ess_w |>
  group_by(country) |>
  summarise(pr = weighted.mean(news_regular, pweight),
            .groups = "drop")

ggplot(pr_cty, aes(x = country, y = pr)) +
  geom_col(fill = "#72B7B2", width = 0.55) +
  labs(y = "Pr(regular news, weighted)", x = "Country",
       title = "Weighted prevalence by country") +
  ylim(0, max(pr_cty$pr) + 0.1) +
  theme_minimal()
```

## 2. Weighted regression: linear probability and logit

```{r svyglm-lpm}
lpm_w <- svyglm(news_regular ~ agea + gender + country,
                design = des, family = gaussian())
tidy(lpm_w)
```

```{r svyglm-logit}
logit_w <- svyglm(news_regular ~ agea + gender + country,
                  design = des, family = quasibinomial())
tidy(logit_w, exponentiate = TRUE)
```

Comparison points for students:

- **Weights + clustering**: `svyglm` gives design-consistent SEs; plain `glm` does not.
- **Quasibinomial** keeps logit link but uses robust variance; estimates mirror survey-weighted MLE when weights are scaled.
- **LPM vs logit under weights**: LPM slopes stay probability-difference interpretation; logit ORs remain multiplicative.

## 3. When to weight (practical guidance)

- Use design/post-strat weights when estimating **population levels** (means, totals, prevalence) or effects that might shift with differential selection.
- In randomized experiments or when modeling **causal effects with ignorable sampling**, weights may be optional; still cluster-robust SEs matter.
- If the research question is **sample-only prediction**, weights can be skipped, but report that scope-of-inference is limited.

## 4. Small worked example: weighted marginal effect of age

```{r weighted-marginal, echo=FALSE}
age_seq <- seq(min(ess_w$agea, na.rm = TRUE), max(ess_w$agea, na.rm = TRUE), length.out = 60)
nd_w <- data.frame(agea = age_seq,
                   gender = "Female",
                   country = "GB")
nd_w$pr <- predict(logit_w, newdata = nd_w, type = "response")

ggplot(nd_w, aes(x = agea, y = pr)) +
  geom_line(color = "#4C78A8", size = 1) +
  labs(y = "Pr(regular news | weighted logit)", x = "Age",
       title = "Weighted predicted probability across age (GB, Female)") +
  theme_minimal()
```

## Practice prompts

1. Recompute weighted country gaps using `anweight` instead of `pweight`. How do the estimates move?
2. Add education to the weighted logit. Do ORs shift relative to the unweighted model in Chapter 5?
3. Estimate a design effect for `news_regular` and discuss how many “effective” observations the design corresponds to.
